---
title: 'Lyons Tran, GGIS 224 Lab 11. Interpolating Temperature Data'
date: "4/19/2024"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

This exercise was modified based on https://geodacenter.github.io/aot-workshop/Part2-AOT.html.

Goals for this lab:
* Deepen the understanding of surface interpolation and the procedure of ordinary kriging. 
* Practice using R's variogram and kriging functions (more at: https://gsp.humboldt.edu/olm/R/04_01_Variograms.html) to perform ordinary kriging for temperature point data.

Go through the steps below and understand the meaning of each script line.
Then, answer the question at the end of this file.

## 1 Environment setup
```{r environment-setup, warning=FALSE, message=FALSE}
library(lubridate) #data wrangling
library(sf)
library(sp) #spatial data wrangling & analysis
library(raster) #spatial raster data wrangling
library(terra) #spatial raster data wrangling
#install.packages("gstat")
library(gstat) #kriging and geostatistics
library(tmap) #modern data visualizations
library(leaflet) #modern data visualizations

# Define your working directory
setwd("C:/Users/lyons/OneDrive/Desktop/ggis224")
```


## 2 Import temperature point data
```{r code chunk for reading the point data}
nodes <- read.csv("Data_Lab11/AOT_node_temps.txt") #Read the file as csv
head(nodes) #Inspect the data: 31 points with temperature values in both Celsius and Fahrenheit degrees.
```


## 3 Convert to spatial data format
Convert the completed node data to spatial object format for plotting and spatial analytics.
```{r code chunk for converting data to spatial format}
node.temps <- nodes #Duplicate 'nodes' to another data file
coordinates(node.temps) <- node.temps[,c("lon", "lat")]
proj4string(node.temps) <- CRS("+init=epsg:4326")
```


## 4 Inspect the converted data
There are 31 sensor points with temperature measurements in both Celsius and Fahrenheit degrees.
```{r code chunk for data inspection}
length(node.temps)
head(node.temps) #avg_temp for Celsius; avg_temp_f for Fahrenheit. 
```


## 5 Plot temperature data
Confirm the success of spatial object transformation by simple plotting.
```{r code chunk for data plotting}
tmap_mode("view")
tm_shape(node.temps) + tm_dots()
```


## 6 Overlay community areas
We here add the Chicago Community Area spatial data to provide additional context for our maps.
The Chicago Community Area shapefile is downloaded from the Chicago Data Portal website: https://data.cityofchicago.org/Facilities-Geographic-Boundaries/Boundaries-Community-Areas-current-/cauq-8yn6. 
```{r code chunk for reading the community area shapefile}
chiCA <- read_sf('Data_Lab11/ChicagoCommunityAreas/geo_export_fcf7548c-1d07-4c42-99a3-156e428257cc.shp')
```

Plot the temperature by sensor points, adding Community Areas as a background.
```{r code chunk for plotting the temperature points and community areas}
tmap_mode("view")
tm_shape(chiCA) + tm_borders() + 
  tm_shape(node.temps) + tm_dots(col="avg_temp_f",size=0.3,title="average temp (F)") 
```


## 7 Interpolate a temperature surface
We will use (semi)variograms to model the distribution of temperature data. We will then generate a grid on top of the Chicago area, and with the appropriate variogram model selected, use kriging to predict a temperature surface.

A variogram is a function that describes the degree of spatial autocorrelation across data. The final model uses the measure of variability between points at various distances. Points nearby are likely to have more similar values, and as distance between points increases, there is less likely to be similar values between points. In this application, we assume that temperature measurements that are further apart will vary more than measurements taken close together (so, a positive autocorrelation).

The variogram clearly has an outlier, though it may not influence our final predicted surface because the variogram will be fitted by a model.
```{r code chunk for plotting the variogram}
tmp.vgm <- variogram(node.temps$avg_temp_f ~ 1, node.temps) #here variable avg_temp_f (temperature measurements in Fahrenheit) is used. 
plot(tmp.vgm)
```


## 8 Semivariogram fitting: using the Spherical model
We will generate a theoretical spherical model to approximate (fit) the experimental semivariogram.
```{r code chunk for fitting the variogram using a spherical model}
tmp.fit.sph<- fit.variogram(tmp.vgm, model=vgm("Sph"))
plot(tmp.vgm, tmp.fit.sph)
```


## 9 Generate grid
Next, we’ll create a grid from the Chicago area. The following function will generate a n-by-n grid from a provided spatial data frame.
```{r code chunk for generating a grid of n cells from a given spatial data frame}
pt2grid <- function(ptframe,n) {
  bb <- bbox(ptframe)  
  ptcrs <- proj4string(ptframe)  
  xrange <- abs(bb[1,1] - bb[1,2])  
  yrange <- abs(bb[2,1] - bb[2,2])  
  cs <- c(xrange/n,yrange/n)  
  cc <- bb[,1] + (cs/2)  
  dc <- c(n,n)  
  x1 <- GridTopology(cellcentre.offset=cc,cellsize=cs,cells.dim=dc)  
  x2 <- SpatialGrid(grid=x1,proj4string=CRS(ptcrs))
  return(x2)
}
```

Let’s first generate a grid of 30 by 30 cells using the Chicago Community Area extent. Plot the grid for exploration.
```{r code chunk for generating a 30x30 grid using the Chicago Community Area as the spatial data frame}
chiCA = as_Spatial(chiCA) #converting chiCA (sf) to a spatial data frame object (needed for executing the pt2grid function)
chi.grid <- pt2grid(chiCA,30)
plot(chi.grid)
```

To get an even finer resolution, we generate a finer resolution of grid of 100 by 100 cells. Let's use the finer grid for the following analysis.
```{r code chunk for generating a 100x100 grid using the Chicago Community Area as the spatial data frame}
chi.grid <- pt2grid(chiCA,100)
plot(chi.grid)
```


## 10 Prepare data for kriging
First, we make sure that all our data is in the same projection.
```{r code chunk for transforming projections}
projection(chi.grid) <- crs("+init=epsg:4326")  
projection(node.temps) <-  crs("+init=epsg:4326")
projection(chiCA) <- crs("+init=epsg:4326")
```

Krige the data using the spherical model.
```{r code chunk for kriging the data using the fitted spherical semivariogram model}
temp.kriged <- krige(node.temps$avg_temp_f ~ 1, node.temps, chi.grid, model = tmp.fit.sph) #Again, variable avg_temp_f is used for kriging. 
plot(temp.kriged)
```

Convert the files to SpatRaster and SpatVector files, clip to Chicago boundaries, and plot to confirm everything looks good.
```{r code chunk for clipping the kriging result to Chicago extent}
chiCA.spat <- terra::vect(chiCA) #convert so we can use in Terra
temp.kriged.spat <- terra::rast(temp.kriged) #convert so we can use in Terra
chi.temp.kriged <- terra::crop(temp.kriged.spat, chiCA.spat, mask=TRUE) #use chiCA as a mask so we only see temperatures within the Chicago boundary
plot(chi.temp.kriged$var1.pred)
plot(node.temps, add = TRUE)
plot(chiCA, add = TRUE)
```

Map the kriged Chicago-area surface. 

```{r code chunk for mapping the kriging result on an interactive view}
tmap_mode("view")
tm_shape(chiCA) + tm_borders() + 
  tm_shape(node.temps) + tm_dots(size=0.01) +
  tm_shape(chi.temp.kriged$var1.pred) + tm_raster("var1.pred", style = "jenks", title = "Temperature (F)", palette = "BuPu") +
  tm_legend(position = c("left", "bottom")) #var1.pred is the ordinary kriging predictions, and #var1.var is the ordinary kriging variance
```

## Question 
Now it is your turn. Try to krige the temperature values (this time, in Celsius degree) using a exponential fitting model, and map both of the predicted surface and its variance. 
Hint: to fit the semivariogram by an exponential model, change the argument of model in fit.variogram to "Exp". 
Also see more: https://search.r-project.org/CRAN/refmans/gstat/html/fit.variogram.html. 
```{r code chunk for kriging temperature in C using an exponential model}
#Write your code chunk below


temperature.krigeInCelsius = variogram(node.temps$avg_temp ~ 1, node.temps)#avg_temp is in celsius
tempKC.fit.exp = fit.variogram(temperature.krigeInCelsius, model = vgm("Exp"))#referred to website linked at the beginning(https://geodacenter.github.io/aot-workshop/Part2-AOT.html) for this and lab example 
plot(temperature.krigeInCelsius)
plot(temperature.krigeInCelsius, tempKC.fit.exp)


temp.kriged.C = krige(avg_temp ~ 1, node.temps, chi.grid, model = tempKC.fit.exp)
temp.kriged.C.spat = terra::rast(temp.kriged.C)
chi.temp.kriged.C <- terra::crop(temp.kriged.C.spat, chiCA.spat, mask=TRUE)
plot(chi.temp.kriged.C$var1.pred)
plot(node.temps, add = TRUE)
plot(chiCA, add = TRUE)

tmap_mode("view")
tm_shape(chiCA) + tm_borders() + 
  tm_shape(node.temps) + tm_dots(size=0.01) +
  tm_shape(chi.temp.kriged.C$var1.pred) + tm_raster("var1.pred", style = "jenks", title = "Temperature in Celsius", palette = "BuPu") +
  tm_shape(chi.temp.kriged.C$var1.var) + tm_raster("var1.var", style = "jenks", title = "Plotting Variance", palette = "BuPu") + #used var1.var as noted in comment earlier
  tm_legend(position = c("left", "bottom"))

```
.